Fail Faster

### 2025-03-14 paper/paper.tex
cut at rng section

The core of any PBT generator library is a (pseudo) random number generator.
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm.

Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
an impediment towards better bugfinding ability \cn \jwc{see comment for cites}.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

Significantly faster RNGs like the Lehmer RNG \cn exist like, and using them can
speed up generation tremendously --- in microbenchmarks, a C implementation of
the lehmer generator runs 4x faster than a C implementation of splitmix.
Similarly, PBT libraries could even eschew the requirement that a source of
entropy pass statistical tests\cn.  This is common practice in testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn.  Bumping a pointer and reading from a buffer (which can be pipelined
trivially) will always be significantly faster than any RNG with data-dependent
arithmetic instructions.

In the context of this paper, however, demonstrating the benefits of any of the
above options is tricky.  Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property.  In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.

>>>>

The core of any PBT generator library is a (pseudo) random number generator.
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm \cn.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

The RNG sits at the heart of the generation hot path.
Even basic generators like generating a single \texttt{int} or \texttt{float} may sample from the RNG multiple times,
and generator combinators like \texttt{list} generate many \texttt{int}s.
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
we argue that existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

For example, significantly faster RNG algorithms than SplitMix or JSF exist,
such as the Lehmer RNG \cn exist. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
% https://github.com/lemire/testingRNG/tree/master
Moreover, PBT libraries could even consider eschewing the requirement that a source of
entropy pass statistical tests like BigCrush\cn.  This is common practice in other areas of testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
pointer and reading from memory (which can be pipelined trivially) will always
be faster on modern CPUs than any RNG with data-dependent arithmetic
instructions.

In the context of this paper, however, demonstrating the benefits of any of the
above faster options is tricky. Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property.  In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.
